Holomorphic field theories and higher algebra

### 2025-01-10 draft.tex
the seal is broken

@@

\usepackage{parskip}
\setlength{\parindent}{18pt}
\setlength{\parindent}{0cm}
\setcounter{tocdepth}{1}

@@

\def\g{{\mathfrak g}}
\def\U{{\rm U}}
\def\mc{\mathcal}
\def\mcol{\, | \,}
\def\ot{\otimes}
\def\Disj{\operatorname{Disj}}
\def\Open{\operatorname{Open}}
\def\Vect{\operatorname{Vect}}
\def\jou{{\mathtt{j}}}
\def\bgamma{{\mathbf{\gamma}}}
\def\bbeta{{\mathbf{\beta}}}
\def\CDO{{\cC\cD}}
\def\SU{{\rm SU}}
\def\sfa{\mathsf{a}}
\def\del{\partial}
\def\til{\Tilde}
\def\Sol{\text{Sol}}
\def\Spec{\text{Spec}}
\def\C{{\mathbb{C}}}

@@

%%% TIKZ STUFF
\usepackage{tikz,tikz-cd}
\usetikzlibrary{arrows,shapes}
\usetikzlibrary{trees}
\usetikzlibrary{matrix,arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{calc,through}
\usetikzlibrary{decorations.pathreplacing}
\usepackage{pgffor}

%%Feynman diagrams
\usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{decorations.markings}
\tikzset{
	% >=stealth', %%  Uncomment for more conventional arrows
    vector/.style={decorate, decoration={snake}, draw},
	provector/.style={decorate, decoration={snake,amplitude=2.5pt}, draw},
	antivector/.style={decorate, decoration={snake,amplitude=-2.5pt}, draw},
    fermion/.style={draw=black, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw=black]{>}}}},
    fermionbar/.style={draw=black, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw=black]{<}}}},
    fermionnoarrow/.style={draw=black},
    gluon/.style={decorate, draw=black,
        decoration={coil,amplitude=4pt, segment length=5pt}},
    scalar/.style={dashed,draw=black, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw=black]{>}}}},
    scalarbar/.style={dashed,draw=black, postaction={decorate},
        dwecoration={markings,mark=at position .55 with {\arrow[draw=black]{<}}}},
    scalarnoarrow/.style={dashed,draw=black},
    electron/.style={draw=black, postaction={decorate}/,
        decoration={markings,mark=at position .55 with {\arrow[draw=black]{>}}}},
	bigvector/.style={decorate, decoration={snake,amplitude=4pt}, draw},
}

@@

As a more mundane example, let $\RR\Hol(X,Y)$ denote the derived moduli space of holomorphic maps from some complex manifold $X$ to another complex manifold $Y$.
(For $Y = V$ a vector space, a ``derived'' enhancement is familiar as the higher dimensional cohomology $H^*(X, \cO) \otimes V$.)
In derived geometry one can take a ``shifted cotangent bundle'' $T^*[-1] \RR\Hol(X,V)$,
and this derived moduli space encodes solutions to
a field theory known as the $\beta\gamma$ system with target~$V$.\footnote{The name is silly: it is due to the original notation people used.}
Every question one asks about the classical physics of the $\beta\gamma$ theory translates into a question about this moduli space.
Other classes of examples are afforded by holomorphic variants of the AKSZ $\sigma$-models (such as the Poisson $\sigma$-model that appears in Kontsevich's construction of deformation quantization); 
the relevant moduli space being the derived mapping space between complex manifolds.

@@

It is natural to wonder how the myriad tools and computations deployed on a supersymmetric theory carry over to this twist 
and, conversely, whether information can flow back from a holomorphically twisted theory to the original supersymmetric theory, 
much as TFTs have enriched the understanding of supersymmetric theories.
In short, a holomorphic twist contains the ``holomorphic sector'' of the untwisted theory just as a topological twist contains its ``topological sector.''
To aid translation between mathematics and physics, 
it may be helpful to know that the role of derived geometry on the mathematical side is mirrored by the use of the BRST/BV formalism on the physics side;
they are closely related and serve similar purposes.

As a final motivational remark, we note in one complex dimension (i.e., working on Riemann surfaces), 
holomorphic field theories appear as chiral conformal field theories (CFTs) on Riemann surfaces. 
The study of CFTs--- whose manifestations in mathematics include vertex algebras, affine Lie algebras, and loop groups --- have had a strong impact in mathematics,
so it is natural to explore higher-dimensional analogs.
So far, most mathematical effort in this direction has focused on (essentially) classical aspects of gauge theory,
but the quantum aspects are profoundly interesting.
We expect that these holomorphic examples will be particularly tractable and beautiful but also useful in offering insights even for non-holomorphic theories.

@@

Over the last few decades there has been a vigorous dialogue between complex geometry and theoretical physics,
around topics like mirror symmetry, enumerative algebraic geometry, moduli of Calabi-Yau manifolds, the geometric Langlands program, and integrable systems, among others.
Covering all these topics would be difficult.
Our survey's focus will be narrow:
it is upon higher dimensional versions of the relationship between geometry and algebraic structures like Lie algebras, Poisson algebras, and vertex algebras.
We will emphasize analogies to TFTs (such as the Gerstenhaber algebras appearing in mirror symmetry) and CFTs (such as the affine Lie algebras relating to moduli of $G$-bundles on Riemann surfaces),
but our primary interest is on complex manifolds above dimension one.
Experience suggests that higher dimensions require higher algebra,
i.e., algebraic methods that involve cochain complexes.
Thus differential graded (dg) algebras and categories,
as well as fellow travelers like derived geometry, appear.

A community has formed, involving both mathematicians and physicists, to explore and exploit these relationships,
and it continues to grow as there is a lot to explore here.
Examples of efforts in this direction include \cite{PTVV}, \cite{FHK,HenKap}, \cite{CLbcov, LiVertex}, \cite{CGholography,BGKWY}, \cite{CostelloPaquette}, \cite{BBBDN}, and \cite{AlfonsiYoungHC},
where our own background and interests (and unfortunately, our ignorance) have shaped this list.

@@

Beyond his writing, however, we have also benefited throughout our careers from the attitudes and atmosphere that his persona modeled and spread among mathematicians:
a generosity with ideas but also poetry and literature, an exquisite care and sensitivity to organizing and explaining mathematics (and much else),
and a kindness in conversation, where he brought his whole attention to you.
While a graduate student at Northwestern, 
OG had the unbelievable luck to overlap with Manin and thus to take several courses and have many conversations with him, 
getting to witness his personification of the Platonic idea of mathematician as intellectual in the broadest sense.
We both spent extensive time, for example, at the Max Planck Institute for Mathematics,
which is suffused with his spirit.
We wish we had the chance to discuss the ideas from this survey with Manin;
we offer it as a small gift in gratitude for all he shared with us.

@@

We begin by sketching what a classical field theory is, somewhat impressionistically, and then offer some concrete examples.

In physics, a classical field theory is usually described in terms of a few pieces of data:
\begin{itemize}
\item a manifold $M$ that plays the role of ``spacetime,''
\item a space of fields, such as sections of some fiber bundle $F \to M$ or connections on a principal bundle on $M$, and
\item a system of partial differential equations (PDE) for the fields, known as the equations of motion.
\end{itemize}
These equations of motion are variational: 
they are the Euler-Lagrange equations of an action functional on the space of fields,
so that solutions to these PDE are the critical points, in a sense, of some function.
In practice an action functional is a formal expression, 
written as an integral over $M$ of some polydifferential operator on the fields that takes values in densities of~$M$.
A few examples will illuminate this sketch.

\begin{eg}[Massless scalar field theory]
\label{eg: massless scalar}
Let $M$ be a smooth manifold and equip it with a Riemannian metric~$g$.
Let $\Delta_g$ denote the associated Laplace-Beltrami operator, 
and let $\d{\rm vol}_g$ denote the associated volume form.
Let the space of fields be the smooth functions $C^\infty(M)$.
A smooth function $\phi$ is called, in physics, a {\em scalar} field.
The action functional
\[
S(\phi) = \int_M \phi \Delta_g \phi \, \d{\rm vol}_g
\]
corresponds to the equation of motion $\Delta_g \phi = 0$. 
Thus this scalar field theory has the moduli space of {\em harmonic functions} as its space of solutions.
\end{eg}

@@

We now describe an example of a theory whose space of solutions is a moduli space of interest in geometric topology.

@@

For a systematic introduction to classical field theory, 
in a style amenable to a complex geometer,
we suggest \cite{DeligneFreed}.
Recently there is an effort to revisit field theory with insights from derived geometry.
See \cite{AlfonsiYoung} for motivation and overview, 
but for a detailed development, see \cite{Steffens23, Steffens24}.
In Section~\ref{sec: pert dfn} we offer a sharp definition well-suited to perturbative field theory 
(i.e., studying formal deformations of a fixed solution to the equations of motion),
which we will use to produce examples of higher algebras.

\subsection{More examples}
\label{sec: ex of theories}

@@

>>>>

\setlength\epigraphwidth{.8\textwidth}

@@

\usepackage{parskip}
\setlength{\parindent}{18pt}
\setlength{\parindent}{0cm}
\setcounter{tocdepth}{2}

@@

@@

@@

As a more mundane example, let $\RR\Hol(X,Y)$ denote the derived moduli space of holomorphic maps from some complex manifold $X$ to another complex manifold $Y$.
For $Y = V$ a complex vector space, this ``derived'' enhancement is simply the entire  cohomology $H^*(X, \cO) \otimes V$, and not just the global functions $H^0(X,\cO) \otimes V$.
There are holomorphic theories whose derived moduli spaces are encoded by cohomology groups of holomorphic vector bundles.
(A physicist might view them as analogs of scalar field theories.)

@@

It is natural to wonder how the myriad tools and computations deployed on a supersymmetric theory carry over to this holomorphic twist 
and, conversely, whether information can flow back from a holomorphically twisted theory to the original supersymmetric theory, 
much as TFTs have enriched physicists' understanding of supersymmetric theories.
In short, a holomorphic twist contains the ``holomorphic sector'' of the untwisted theory just as a topological twist contains its ``topological sector.''
To aid translation between mathematics and physics, 
it may be helpful to know that the role of derived geometry on the mathematical side is mirrored by the use of the BRST/BV formalism on the physics side;
they are closely related and serve similar purposes.

As a final motivational remark, we note that in one complex dimension (i.e., working on Riemann surfaces), 
holomorphic field theories appear as chiral conformal field theories (CFTs) on Riemann surfaces. 
The study of CFTs --- whose manifestations in mathematics include vertex algebras, affine Lie algebras, nonabelian theta functions, and loop groups --- have had a strong impact in mathematics,
so it is natural to explore higher-dimensional analogs.
So far, most mathematical effort in this direction has focused on (essentially) classical aspects of gauge theory,
but the quantum aspects are profoundly interesting.
We expect that these holomorphic examples will be particularly tractable and beautiful but also useful in offering insights even for non-holomorphic theories.

@@

Over the last few decades there has been a vigorous dialogue between complex geometry and theoretical physics,
around topics like mirror symmetry, enumerative algebraic geometry, moduli of Calabi-Yau manifolds, the geometric Langlands program, and integrable systems, among others.
Covering all these topics would be difficult.
Our survey's focus will be narrow:
it is upon higher dimensional versions of the relationship between complex geometry and algebraic structures like Lie algebras, Poisson algebras, and vertex algebras.
We will emphasize analogies to TFTs (such as the Gerstenhaber algebras appearing in mirror symmetry) and CFTs (such as the affine Lie algebras relating to moduli of $G$-bundles on Riemann surfaces),
but our primary interest is on complex manifolds above dimension one.
Experience suggests that higher dimensions require higher algebra,
i.e., algebraic methods that involve cochain complexes.
Thus differential graded (dg) algebras and categories,
as well as fellow travelers like derived geometry, appear.

A community has formed, involving both mathematicians and physicists, to explore and exploit these relationships,
and it continues to grow as there is a lot to explore.
Examples of efforts in this direction include \cite{PTVV}, \cite{FHK,HenKap}, \cite{CLbcov, LiVertex}, \cite{CGholography,BGKWY}, \cite{CostelloPaquette}, \cite{BBBDN}, and \cite{AlfonsiYoungHC},
where our own background and interests (and unfortunately, our ignorance) have shaped this list.

@@

Beyond his writing, however, we have also benefited throughout our careers from the attitudes and atmosphere that his persona modeled and spread among mathematicians:
a generosity with mathematical ideas but also the sharing of poetry and literature, 
an exquisite care and sensitivity to organizing and explaining mathematics (and much else),
and a kindness in conversation, where he brought his whole attention to you.
While a graduate student at Northwestern, 
OG had the unbelievable luck to overlap with Manin and thus to take several courses and have many conversations with him, 
getting to witness his personification of the Platonic idea of mathematician as intellectual in the broadest sense.
We also both spent extensive time, for example, at the Max Planck Institute for Mathematics,
which is suffused with his spirit.
We wish we had the chance to discuss the ideas from this survey with Manin;
we offer it as a small gift in gratitude for all he shared with us.

@@

We begin by explaining the basic framework of classical field theory, with a few examples.
For a mathematician, the key point might be that physics suggests certain systems of equations are particularly important and studies their spaces of solutions, 
often unearthing aspects that might not catch a mathematician's eye.
In other words, physics can be a source of interesting moduli spaces.

We then offer a quick overview of derived geometry, aiming to motivate and not to give a systematic treatment.
For this survey, comfort with basic homological algebra and sheaf cohomology will suffice, 
so we will explain how our terminology relates to those notions.
In

\subsection{A first pass}

In physics, a classical field theory is usually described in terms of a few pieces of data:
\begin{itemize}
\item a manifold $M$ that plays the role of ``spacetime,''
\item a space of ``fields,'' typically the sections of some fiber bundle $F \to M$ or connections on a principal bundle on $M$, and
\item a system of partial differential equations (PDE) for the fields, known as the ``equations of motion.''
\end{itemize}
These equations of motion are variational: 
the solutions to these PDE are the critical points, in a sense, of some function,
called an ``action functional.''
(Variational PDE are often called Euler-Lagrange equations.)
In practice an action functional is a formal expression, 
written as an integral over $M$ of some polydifferential operator on the fields that takes values in densities of~$M$.

A few examples will illuminate this sketch.
We begin by giving two examples that depend on Riemannian geometry, 
then give an example of a topological field theory,
and finally give an example of a holomorphic field theory.
All the equations that show up are well-known to geometers.

\begin{eg}[Massless scalar field theory]
\label{eg: massless scalar}
Let $M$ be a smooth manifold and equip it with a Riemannian metric~$g$.
Let $\Delta_g$ denote the associated Laplace-Beltrami operator, 
and let $\d{\rm vol}_g$ denote the associated volume form.
In the case $M = \RR^n$ with the Euclidean inner product, the Laplace-Beltrami operator is the differential operator
\[
\Delta = \frac{\partial^2}{\partial x_1^2} + \cdots + \frac{\partial^2}{\partial x_n^2}.
\]
Let the space of fields be the smooth functions $C^\infty(M)$.
A smooth function $\phi$ is called, in physics, a {\em scalar} field.
The action functional
\[
S(\phi) = \int_M \phi \Delta_g \phi \, \d{\rm vol}_g
\]
corresponds to the equation of motion $\Delta_g \phi = 0$. 
Thus this scalar field theory has the moduli space of {\em harmonic functions} as its space of solutions.
\end{eg}

@@

We now describe an example of a theory whose space of solutions is a moduli space of interest in geometric topology: character varieties.

@@

For a systematic introduction to classical field theory, 
in a style amenable to a complex geometer,
we suggest \cite{DeligneFreed}.
In the appendix we describe many more examples of holomorphic field theories and describe their spaces of solutions using the language of derived geometry, which we discuss next.

\subsection{A dose of derived geometry}
\label{sec: DAG}

Recently there is an effort to revisit field theory with insights from derived geometry.
See \cite{AlfonsiYoung} for motivation and overview, 
but for a detailed development, see \cite{Steffens23, Steffens24}.
In Section~\ref{sec: pert dfn} we offer a sharp definition well-suited to perturbative field theory 
(i.e., studying formal deformations of a fixed solution to the equations of motion),
which we will use to produce examples of higher algebras.

@@

\section*{Appendix: A zoo of holomorphic field theories}
\label{sec: ex of theories}
